Formula 1 2022 Regulation Changes

### 2024-04-23 paper/paper.qmd
added conclusion and more references

Despite these debates, preliminary findings suggest that the 2022 regulations have had a profound impact on Formula 1. Races have generally become more competitive, with a wider array of teams and drivers finding themselves in contention for podium finishes. The budget cap has also initiated a more strategic allocation of resources among teams, potentially leveling the competitive field over the longer term. The paper's estimand is whether the 2022 Formula 1 regulations led to a significant increase in lap time in comparison to the previous time among the top drivers across different types of circuits but also increased the number of successful overtakes during races. We aim to quantify the impact of the 2022 Formula 1 regulatory changes, providing a clear basis for evaluating their success in enhancing the sport's competitive balance and the quality of racing.

@@

The f1dataR [@f1dataR] package in R is designed specifically for Formula 1 enthusiasts and data analysts who wish to examine the statistical and historical aspects of Formula 1 racing. This package provides a comprehensive setf functions that facilitate the extraction and manipulation of Formula 1 data directly from the Ergast API, which is a widely recognized source for F1 statistics. With f1dataR, users can easily access detailed race results, driver standings, constructor details, and much more, across the history of F1 competitions.

@@

To analyze the impact of track characteristics on Formula 1 car performance, specifically comparing the lap times and speeds at the Saudi Arabia Jeddah Corniche Circuit (a top-speed track) and the Netherlands' Circuit Zandvoort (a high-downforce track) as seen in @tbl-circuits for the 2021 and 2022 seasons, a structured approach will be employed. Initially, data will be gathered using the f1dataR package, which extracts detailed information about lap times and speeds from the Ergast API. This dataset will include drivers' performances during both qualifying and race sessions.

The analysis will begin with descriptive statistics to outline the basic trends in the ranges of lap times and top speeds for each session across both tracks and years. This provides a preliminary view of the data and helps identify any outliers or anomalies. For deeper insights, a comparative analysis will be conducted between the two tracks and between the two years. This involves comparing average speeds and lap times to determine how the distinct characteristics of each track influence performance and how performance metrics have evolved from one year to the next.

To enhance the method of analysis for comparing lap times and speeds at the Jeddah Corniche Circuit and Circuit Zandvoort over the 2021 and 2022 Formula 1 seasons, we will focus on three top drivers: Max Verstappen, Lewis Hamilton, and Charles Leclerc. Including these specific drivers in the study adds a layer of depth by enabling comparisons not just between tracks and years, but also among leading competitors known for their distinctive driving styles and performance under varying circuit conditions.

@@

variable_data <- data.frame( 
  CircuitName = c("Saudi Arabia", 
               "Netherlands"), 
  TrackType = c("Top Speed",
                  "High Downforce"), 
  TyreWear = c("High", 
              "Low")
)
variable_data |>
 slice(1:10) |>
  kable(digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```

@@

@fig-maxdutch compares lap times achieved by Max Verstappen in the Netherlands for two consecutive Formula 1 seasons, 2021 and 2022. Each point on the graph represents the lap time for a specific lap number during the race. From the pattern of the data points, it can be observed that the majority of lap times for both years are closely clustered at the lower end of the lap time scale, which indicates consistent performance across the majority of the laps.

In comparing the two seasons, the 2022 lap times appear to be more volatile with higher spikes, suggesting that there were more instances where Verstappen had to slow down significantly. This might be reflective of strategic elements such as pit stop timing or could be indicative of the effects of the new regulations on race strategy and car performance. Additionally, the 2022 data points generally seem to trend slightly higher than those of 2021, which imply that the lap times were slower in 2022 on average.

@@

@fig-lewisdutch shows Lewis Hamilton's lap times in the Netherlands for the 2021 and 2022 Formula 1 seasons. Each dot represents a lap time for a corresponding lap number.

Looking at the 2021 season, represented by the purple dots, we see a consistent baseline of lap times with occasional spikes. These spikes likely indicate laps where Hamilton pitted, encountered traffic, or reduced speed for safety cars or yellow flags. The baseline lap time is quite stable, suggesting consistent performance throughout the race when unaffected by external factors. The 2022 season, shown in yellow, follows a similar pattern of consistency but with fewer and less pronounced spikes. Moreover, the baseline lap times in 2022 appear to be consistently lower than in 2021. This could suggest an improved car performance or a better adaptation to the track conditions.

@@

Comparing the two seasons, the 2022 green dots display a trend of marginally lower lap times than in 2021, suggesting an improved performance in the 2022 season. This could be due to advancements in car performance, a better understanding of the track, tire strategies, or other race conditions.

It's noteworthy that in both seasons, the spike patterns are somewhat similar, indicating that the race circumstances that caused these laps to be slower were likely regular race events like pit stops. The spikes are also comparable in magnitude between the two years, which suggests that the time spent on these slower laps did not change dramatically from year to year.

@@

The top graph shows a general speed comparison throughout a single lap. Both lines (2021, 2022) display similar shapes, which means that the speed profiles for the lap are fairly consistent from year to year, suggesting that the general flow and demands of the track haven't changed dramatically. However, the 2022 line largely sits above the 2021 line, indicating that the drivers speeds were generally higher during the qualifying lap in the 2022 season. This could be attributed to several factors including changes in regulations leading to faster cars, or improved engine performance.

The bottom graph zeroes in on the speeds at the final corner of the lap. It's noticeable that the 2022 speed starts higher but then drops below the 2021 speed as the corner progresses. This suggests in terms of straight-line speed, the 2022 cars benefit from a reduction in drag due to sleeker aerodynamic profiles and simpler wing designs. This reduction allows the cars to achieve higher top speeds, as drag is a key limiting factor on straights. The graph reflecting the drivers speed during a qualifying lap in Jeddah corroborates this, with the 2022 car showing a generally higher velocity compared to the 2021 car, suggesting a possible increase in aerodynamic efficiency.

@@

The 2022 regulations introduced a shift back to ground effect aerodynamics, a method that relies on the car’s underbody design to generate a large portion of its downforce. This allows for more stable aerodynamics when following another car closely, which ideally enhances overtaking opportunities and racing quality. However, these changes also mean that the cars generate less downforce overall compared to the high-rake aerodynamics used in previous years [@f1_changes]. Less downforce generally leads to slower speeds in high-speed corners, which can contribute to slower lap times overall.

Another significant factor is the increase in the minimum weight of the cars. The 2022 regulations raised the minimum weight from 752 kg to 798 kg, largely due to the introduction of heavier and safer chassis requirements and other components like bigger wheels (18 inches up from 13 inches) [@f1_changes]. Heavier cars require more energy to accelerate and decelerate, which naturally results in slower lap times, especially evident in tighter, more technical circuits where agility and quick changes of direction are crucial.

The switch to 18-inch wheels from the previous 13-inch also impacts the car's performance. The larger rims come with a lower-profile tire, which behaves differently compared to the high-profile tires used earlier. These changes affect the suspension geometry and dynamics, as well as how the tire copes with heat and wear during a race. While designed to be more durable, the new tires may not provide the same level of grip and responsiveness as the previous designs, potentially leading to slower lap times. 2022 F1 cars are engineered with a different set of priorities and constraints, focusing on enhancing the racing experience and safety [@f1_changes]. While this might mean they are initially slower in terms of lap times, they offer improved performance in terms of race dynamics, such as in duels and overtakes.

@@

The 2022 Formula 1 regulations demonstrate how precise, targeted regulatory changes can substantially enhance the dynamics of competition within sports, offering a rich lesson in the effectiveness of thoughtful rule-making. These regulations primarily focused on redesigning the aerodynamics of the vehicles, aiming to mitigate the 'dirty air' effect—turbulent air left by a leading car that significantly reduces the aerodynamic efficiency of following cars. By introducing cars with ground effect designs that rely less on aerodynamic appendages susceptible to turbulence, the regulations reduced the aerodynamic wake and thereby allowed trailing cars to follow more closely without significant loss of performance.

The impact of these changes on the sport is observable through a marked increase in the number and quality of overtaking maneuvers across various circuits, especially in high-speed corners where following closely used to be particularly challenging. Detailed analysis of race data shows not only a quantitative increase in overtakes but also qualitative improvements in race tactics, as drivers now exploit new overtaking opportunities that previously did not exist. This has led to races that are less predictable and more dynamic, contributing to a more engaging viewer experience and broader fan engagement.

@@

The 2022 Formula 1 regulations bring into focus the economic and environmental dimensions of sports regulation, particularly in how they adapt to contemporary sustainability challenges. The introduction of a cost cap is a pivotal development; it aims to equalize the competitive environment by limiting the amount teams can spend throughout a season. This economic mechanism is designed not only to curb the escalating expenditures that often see wealthier teams pull far ahead but also to foster a more strategically diverse field where innovation must be achieved within fixed financial boundaries. This could lead to smarter, more efficient use of resources, which is essential for the sport's long-term sustainability.

@@

>>>>

Despite these debates, preliminary findings suggest that the 2022 regulations have had a profound impact on Formula 1. Races have generally become more competitive, with a wider array of teams and drivers finding themselves in contention for podium finishes. The budget cap has also initiated a more strategic allocation of resources among teams, potentially leveling the competitive field over the longer term. The paper's estimand is whether the 2022 Formula 1 regulations led to a significant increase in lap time and overall top speed in comparison to the previous time among the top drivers across different types of circuits but also increased the number of successful overtakes during races. We aim to quantify the impact of the 2022 Formula 1 regulatory changes, providing a clear basis for evaluating their success in enhancing the sport's competitive balance and the quality of racing.

@@

The f1dataR [@f1dataR] package in R is designed specifically for Formula 1 enthusiasts and data analysts who wish to examine the statistical and historical aspects of Formula 1 racing. This package provides a comprehensive set of functions that facilitate the extraction and manipulation of Formula 1 data directly from the Ergast API, which is a widely recognized source for F1 statistics. With f1dataR, users can easily access detailed race results, driver standings, constructor details, and much more, across the history of F1 competitions.

@@

To analyze the impact of the F1 Regulations, we are going to be comparing the lap times and speeds at the Saudi Arabia Jeddah Corniche Circuit (a top-speed track) and the Netherlands' Circuit Zandvoort (a high-downforce track) as seen in @tbl-circuits for the 2021 and 2022 seasons, a structured approach will be employed. Initially, data will be gathered using the f1dataR package, which extracts detailed information about lap times and speeds from the Ergast API. This dataset will include drivers' performances during both qualifying and race sessions.

The analysis will be conducted between the two tracks and between the two years. This involves comparing average speeds and lap times to determine how the distinct characteristics of each track influence performance and how performance metrics have evolved from one year to the next. To enhance the method of analysis for comparing lap times and speeds at the Jeddah Corniche Circuit and Circuit Zandvoort over the 2021 and 2022 Formula 1 seasons, we will focus on three top drivers: Max Verstappen, Lewis Hamilton, and Charles Leclerc. Including these specific drivers in the study adds a layer of depth by enabling comparisons not just between tracks and years, but also among leading competitors known for their distinctive driving styles and performance under varying circuit conditions.

@@

variable_data <- data.frame( 
  CircuitName = c("Saudi Arabia Jeddah Corniche Circuit", 
               "Netherlands' Circuit Zandvoort"), 
  TrackType = c("Top Speed",
                  "High Downforce"), 
  TyreWear = c("High", 
              "Low")
)
variable_data |>
 slice(1:10) |>
  kable(digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```

@@

@fig-maxdutch compares lap times achieved by Max Verstappen in the Netherlands for two consecutive Formula 1 seasons, 2021 and 2022. From the pattern of the data points, it can be observed that the majority of lap times for both years are closely clustered at the lower end of the lap time scale, which indicates consistent performance across the majority of the laps.

In comparing the two seasons, the 2022 lap times appear to be more volatile with higher spikes, suggesting that there were more instances where Verstappen had to slow down significantly. This might be reflective of the effects of the new regulations on race strategy and car performance. Additionally, the 2022 data points generally seem to trend slightly higher than those of 2021, which imply that the lap times were slower in 2022 on average.

@@

@fig-lewisdutch shows Lewis Hamilton's lap times in the Netherlands for the 2021 and 2022 Formula 1 seasons. Looking at the 2021 season, represented by the purple dots, we see a consistent baseline of lap times with occasional spikes. The baseline lap time is quite stable, suggesting consistent performance throughout the race when unaffected by external factors. The 2022 season, shown in yellow, follows a similar pattern of consistency but with fewer and less pronounced spikes. Moreover, the baseline lap times in 2022 appear to be consistently higher than in 2021. This implies that laptimes were also slower in 2022.

@@

Comparing the two seasons, the 2022 green dots display a trend of marginally lower lap times than in 2021, suggesting an improved performance in the 2022 season. It's noteworthy that in both seasons, the spike patterns are somewhat similar, indicating that the race circumstances that caused these laps to be slower were likely regular race events like pit stops. The spikes are also comparable in magnitude between the two years, which suggests that the time spent on these slower laps did not change dramatically from year to year.

@@

The top graph shows a general speed comparison throughout a single lap. Both lines (2021, 2022) display similar shapes, which means that the speed profiles for the lap are fairly consistent from year to year, suggesting that the general flow and demands of the track haven't changed dramatically. However, the 2022 line largely sits above the 2021 line, indicating that the drivers speeds were generally higher during the qualifying lap in the 2022 season.

The bottom graph zeroes in on the speeds at the final corner of the lap which is a very high downforce corner. It's noticeable that the 2022 speed starts higher but then drops below the 2021 speed as the corner progresses. This suggests in terms of straight-line speed, the 2022 cars benefit from a reduction in drag due to sleeker aerodynamic profiles and simpler wing designs. This reduction allows the cars to achieve higher top speeds, as drag is a key limiting factor on straights. The graph reflecting the drivers speed during a qualifying lap in Jeddah corroborates this, with the 2022 car showing a generally higher velocity compared to the 2021 car, suggesting a possible increase in aerodynamic efficiency.

@@

The 2022 regulations introduced a shift back to ground effect aerodynamics, a method that relies on the car’s underbody design to generate a large portion of its downforce. This allows for more stable aerodynamics when following another car closely, which ideally enhances overtaking opportunities and racing quality. However, these changes also mean that the cars generate less downforce overall compared to the high-rake aerodynamics used in previous years [@f1_changes]. Less downforce generally leads to slower speeds in high-speed corners, which can contribute to slower lap times overall which is why we see that Max Verstappen, Lewis Hamilton and Charles Leclerc all had relatively slower laptimes from 2021 to 2022.

Another significant factor is the increase in the minimum weight of the cars. The 2022 regulations raised the minimum weight from 748 kg to 798 kg [@f1_weight], largely due to the introduction of heavier and safer chassis requirements and other components like bigger wheels (18 inches up from 13 inches) [@f1_changes]. Heavier cars require more energy to accelerate and decelerate, which naturally results in slower lap times, especially evident in tighter, more technical circuits where agility and quick changes of direction are crucial.

The switch to 18-inch wheels from the previous 13-inch also impacts the car's performance. The larger rims come with a lower-profile tire, which behaves differently compared to the high-profile tires used earlier. These changes affect the suspension geometry and dynamics, as well as how the tire copes with heat and wear during a race [@f1_changes]. While designed to be more durable, the new tires may not provide the same level of grip and responsiveness as the previous designs, potentially leading to slower lap times. 2022 F1 cars are engineered with a different set of priorities and constraints, focusing on enhancing the racing experience and safety [@f1_changes]. While this might mean they are initially slower in terms of lap times, they offer improved performance in terms of race dynamics, such as in duels and overtakes.

@@

The 2022 Formula 1 regulations demonstrate how precise, targeted regulatory changes can substantially enhance the dynamics of competition within sports. These regulations primarily focused on redesigning the aerodynamics of the vehicles, aiming to mitigate the 'dirty air' effect—turbulent air left by a leading car that significantly reduces the aerodynamic efficiency of following cars. By introducing cars with ground effect designs that rely less on aerodynamic appendages susceptible to turbulence, the regulations reduced the aerodynamic wake and thereby allowed trailing cars to follow more closely without significant loss of performance.

The impact of these changes on the sport is observable through a marked increase in the number and quality of overtaking maneuvers across various circuits, especially in high-speed corners where following closely used to be particularly challenging. @fig-averageovertakes and @fig-overtakes shows not only a quantitative increase in overtakes but also qualitative improvements in race tactics, as drivers now exploit new overtaking opportunities that previously did not exist. This has led to races that are less predictable and more dynamic, contributing to a more engaging viewer experience and broader fan engagement.

@@

The 2022 Formula 1 regulations bring into focus the economic and environmental dimensions of sports regulation, particularly in how they adapt to contemporary sustainability challenges. The introduction of a cost cap is a pivotal development as it aims to equalize the competitive environment by limiting the amount teams can spend throughout a season [@f1_costcap]. This economic mechanism is designed not only to stop the escalating expenditures that often see wealthier teams pull far ahead but also to create a more strategically diverse field where innovation must be achieved within fixed financial boundaries. This could lead to smarter, more efficient use of resources, which is essential for the sport's long-term sustainability.

@@

The 2022 Formula 1 regulations represent a pivotal moment in the sport’s history, aiming to enhance competitiveness and sustainability while addressing both the environmental and economic challenges of modern motorsport. These regulations have introduced significant changes in car design, notably through ground effect aerodynamics, and have implemented a cost cap that levels the playing field, allowing for a more diverse and competitive environment. Preliminary observations indicate that these changes have indeed increased the quality of racing, as evidenced by more frequent and strategic overtaking maneuvers and closer competition among teams.

My analysis began with that the 2022 cars would exhibit faster straight-line speeds due to reduced drag but suffer in cornering speeds due to less downforce. The data collected and analyzed, confirmed this. All drivers showed an increase in top speed from 2021 to 2022, underscoring the effects of aerodynamic simplification. Yet, regardless of the type of corner whether high-speed or low-speed, all tracks examined demonstrated an increase in lap times. This suggests that while the cars gained speed in the straights, they lost substantial time in the corners, leading to overall slower lap times across a full circuit.

Despite these findings, the regulations have positively influenced the sport by enhancing race competitiveness and overtaking, as well as by advancing environmental considerations through the use of sustainable fuels and more stringent aerodynamic rules. As Formula 1 continues to innovate within these new guidelines, it maintains its status as a pinnacle of motorsport technology and strategy.

Looking ahead, further studies are warranted to explore the long-term impacts of these regulations, particularly as the sport approaches another significant regulatory shift in 2026 with new engines, aerodynamics, and fully sustainable fuels. The ongoing challenge for Formula 1 will be to balance technological innovation and competitive racing with environmental sustainability and economic viability, ensuring that the sport remains at the front of global motorsport while adapting to the demands of a changing world.
